New and old ways to chase a rabbit:\\ Differential item functioning detection methods

### 2020-03-02 paper/paper.Rmd
implementing non compensatory models which seems to be going well

@@

# TO DO

- be sure all my directions match. i think i switched everything to be item easiness at some point but not sure if i did it throughout

>>>>

<!-- the literature basically talks separately about anchor selection and DIF methods. much more literature is on DIF methods and less is on anchor selection. I'm basically writing a paper about anchor selection and just calling it DIF methods because that seems like where the action is. I'm taking fixed at that the likelihood ratio test is the way to go (and perhaps that deserves more detail)? -->

@@

### 2020-03-02 paper/paper.tex
implementing non compensatory models which seems to be going well

Contrarily, the usual setup of DIF simulation studies frames DIF as a property of the item. For example, Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) simulate students as belonging to either a reference or focal group. They fix the item easinesses for the reference group, \({b_j}^{\text{ref}}\), to values from a previous study. They set item easinesses for the focal group to \({b_j}^{\text{foc}} = {b_j}^{\text{ref}}\) for items without DIF and \({b_j}^{\text{foc}} = {b_j}^{\text{ref}} + 0.6\) for items with DIF, where 0.6 is the magnitude of DIF in logits. They simulate student ability \({\theta_i}^{\text{ref}} \sim N(0,1)\) for students in the reference group and \({\theta_i}^{\text{foc}} \sim N(-1,1)\) for students in the focal group. They generate item responses according to the Rasch model, which specifies that the probability of student \(i\) responding correctly to item \(j\) is
\begin{align}
    P(y_{ij} = 1 | \theta_i, b_j) = \sigma(\theta_i + b_j)
\end{align}
where \(\sigma(x) = \frac{e^x}{e^x + 1}\) is the standard logistic function.

@@

The anchor point, \(\mu^{\star\text{foc}}\), is selected by maximizing the Gini index. The intuition and assumption is that some items might have bias but the majority do not. Denoting a function that calculates the Gini index from a vector of non-negative elements as \(G(\mathbf{x})\),\\
\[
\mu^{\star\text{foc}} = \mathop\mathrm{arg\,max}_{\mu^\text{foc}} G(|\mu^\text{foc} + \tilde{\mathbf{d}}|)
\]
where \(\mu^\text{foc} \in (-\infty, \infty)\) and \(\mu^\text{foc}\) is added to each element of \(\mathbf{d}\).

@@

Raju's area method measures the amount of DIF by calculating the area between the item characteristic curve, the function that maps ability to probability correct response, for each group (Raju \protect\hyperlink{ref-raju1988area}{1988}):

\begin{align}
    \text{Area Between Curves} = \int |\text{Pr}(y_j = 1| \theta, b_j^{\text{ref}}) - \text{Pr}(y_j = 1| \theta, b_j^{\text{foc}})|
\end{align}

Raju's area method, along with AOAA, have been cited as the two most commonly used IRT-based DIF detection methods (Magis et al. \protect\hyperlink{ref-magis2011generalized}{2011}). We see two weakness of Raju's area method: First, the item characteric curves still need to be linked by anchor items, which the method does not help to identify. Second, the area is unweighted so all values of \(\theta\) matter equally, despite some being much more realistic than others.

>>>>

Contrarily, the usual setup of DIF simulation studies frames DIF as a property of the item. For example, Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) simulate students as belonging to either a reference or focal group. They fix the item easinesses for the reference group, \({b_j}^{\text{ref}}\), to values from a previous study. They set item easinesses for the focal group to \({b_j}^{\text{foc}} = {b_j}^{\text{ref}}\) for items without DIF and \({b_j}^{\text{foc}} = {b_j}^{\text{ref}} + 0.6\) for items with DIF, where 0.6 is the magnitude of DIF in logits. They simulate student ability \({\theta_i}^{\text{ref}} \sim N(0,1)\) for students in the reference group and \({\theta_i}^{\text{foc}} \sim N(-1,1)\) for students in the focal group. They generate item responses according to the Rasch model, which specifies that the probability of student \(i\) responding correctly to item \(j\) is \[P(y_{ij} = 1 | \theta_i, b_j) = \sigma(\theta_i + b_j)\] where \(\sigma(x) = \frac{e^x}{e^x + 1}\) is the standard logistic function.

@@

The anchor point, \(\mu^{\star\text{foc}}\), is selected by maximizing the Gini index. The intuition and assumption is that some items might have bias but the majority do not. Denoting a function that calculates the Gini index from a vector of non-negative elements as \(G(\mathbf{x})\),
\[
\mu^{\star\text{foc}} = \mathop\mathrm{arg\,max}_{\mu^\text{foc}} G(|\mu^\text{foc} + \tilde{\mathbf{d}}|)
\]
where \(\mu^\text{foc} \in (-\infty, \infty)\) and \(\mu^\text{foc}\) is added to each element of \(\mathbf{d}\).

@@

Raju's area method (Raju \protect\hyperlink{ref-raju1988area}{1988}) measures the amount of DIF by calculating the area between the item characteristic curve, the function that maps ability to probability correct response, for each group:
\[
\text{Area Between Curves} = \int |\text{Pr}(y_j = 1| \theta, b_j^{\text{ref}}) - \text{Pr}(y_j = 1| \theta, b_j^{\text{foc}})|
\]
Raju's area method, along with AOAA, have been cited as the two most commonly used IRT-based DIF detection methods (Magis et al. \protect\hyperlink{ref-magis2011generalized}{2011}). We see two weakness of Raju's area method: First, the item characteric curves still need to be linked by anchor items, which the method does not help to identify. Second, the area is unweighted so all values of \(\theta\) matter equally, despite some being much more realistic than others.

### 2020-03-02 paper/paper_1_intro.Rmd
implementing non compensatory models which seems to be going well

@@

Contrarily, the usual setup of DIF simulation studies frames DIF as a property of the item. For example, @kopf2015framework simulate students as belonging to either a reference or focal group. They fix the item easinesses for the reference group, ${b_j}^{\text{ref}}$, to values from a previous study. They set item easinesses for the focal group to ${b_j}^{\text{foc}} = {b_j}^{\text{ref}}$ for items without DIF and ${b_j}^{\text{foc}} = {b_j}^{\text{ref}} + 0.6$ for items with DIF, where 0.6 is the magnitude of DIF in logits. They simulate student ability ${\theta_i}^{\text{ref}} \sim N(0,1)$ for students in the reference group and ${\theta_i}^{\text{foc}} \sim N(-1,1)$ for students in the focal group. They generate item responses according to the Rasch model, which specifies that the probability of student $i$ responding correctly to item $j$ is
\begin{align}
    P(y_{ij} = 1 | \theta_i, b_j) = \sigma(\theta_i + b_j)
\end{align}
where $\sigma(x) = \frac{e^x}{e^x + 1}$ is the standard logistic function.

We find it informative to describe simulation conditions in terms of multidimensional abilities as opposed to item parameters that vary across groups. For example, to translate the @kopf2015framework simulation from the DIF-as-item-property view to the DIF-as-student-property view, item easiness is set to what was previously ${b_j}^{\text{ref}}$ for all students. Student ability is expanded to two dimensions, the target ability dimension and nuisance ability dimension. The target ability is what was previously just ability where ${\theta_i}^{\text{ref}} \sim N(0,1)$ for students in the reference group and ${\theta_i}^{\text{foc}} \sim N(-1,1)$ for students in the focal group. The nuisance ability is set to ${\eta_i}^{\text{ref}} = 0$ for the reference group and ${\eta_i}^{\text{ref}} = -1$ for the focal group. We now need to use a 2PL model where the slope on target ability $a_{1j} = 1$ for all items (consistent with the Rasch model) and the slope on nuisance ability $a_{2j} = 0.6$ for all items with DIF, and $a_{2j} = 0$ otherwise. Again, 0.6 is the magnitude of DIF in logits. According to the two-dimensional 2PL model, the probability student $i$ responding correctly to item $j$ is
\begin{align}
    P(y_{ij} = 1 | \theta_i, \eta_i, a_{1j}, a_{2j}, b_j) = \sigma(a_{1j}\theta_i + a_{2j}\eta_i + b_j).
\end{align}
This translation between views makes explicit that nearly all DIF simulation studies have, perhaps suboptimally, examined the unrealistic scenario in which there is no variation in nuisance ability for students in the same group.

>>>>

<!-- need to specify which multidimensional model im talking about -->

@@

Contrarily, the usual setup of DIF simulation studies frames DIF as a property of the item. For example, @kopf2015framework simulate students as belonging to either a reference or focal group. They fix the item easinesses for the reference group, ${b_j}^{\text{ref}}$, to values from a previous study. They set item easinesses for the focal group to ${b_j}^{\text{foc}} = {b_j}^{\text{ref}}$ for items without DIF and ${b_j}^{\text{foc}} = {b_j}^{\text{ref}} + 0.6$ for items with DIF, where 0.6 is the magnitude of DIF in logits. They simulate student ability ${\theta_i}^{\text{ref}} \sim N(0,1)$ for students in the reference group and ${\theta_i}^{\text{foc}} \sim N(-1,1)$ for students in the focal group. They generate item responses according to the Rasch model, which specifies that the probability of student $i$ responding correctly to item $j$ is $$P(y_{ij} = 1 | \theta_i, b_j) = \sigma(\theta_i + b_j)$$ where $\sigma(x) = \frac{e^x}{e^x + 1}$ is the standard logistic function.

We find it informative to describe simulation conditions in terms of multidimensional abilities as opposed to item parameters that vary across groups. For example, to translate the @kopf2015framework simulation from the DIF-as-item-property view to the DIF-as-student-property view, item easiness is set to what was previously ${b_j}^{\text{ref}}$ for all students. Student ability is expanded to two dimensions, the target ability dimension and nuisance ability dimension. The target ability is what was previously just ability where ${\theta_i}^{\text{ref}} \sim N(0,1)$ for students in the reference group and ${\theta_i}^{\text{foc}} \sim N(-1,1)$ for students in the focal group. The nuisance ability is set to ${\eta_i}^{\text{ref}} = 0$ for the reference group and ${\eta_i}^{\text{ref}} = -1$ for the focal group. We now need to use a 2PL model where the slope on target ability $a_{1j} = 1$ for all items (consistent with the Rasch model) and the slope on nuisance ability $a_{2j} = 0.6$ for all items with DIF, and $a_{2j} = 0$ otherwise. Again, 0.6 is the magnitude of DIF in logits. According to the two-dimensional 2PL model, the probability student $i$ responding correctly to item $j$ is
$$
P(y_{ij} = 1 | \theta_i, \eta_i, a_{1j}, a_{2j}, b_j) = \sigma(a_{1j}\theta_i + a_{2j}\eta_i + b_j).
$$
This translation between views makes explicit that nearly all DIF simulation studies have, perhaps suboptimally, examined the unrealistic scenario in which there is no variation in nuisance ability for students in the same group.

### 2020-03-02 paper/paper_2a_anchor_items.Rmd
implementing non compensatory models which seems to be going well

Psychometricians have long been in search of the perfect DIF detection method. In this section, we summarize existing DIF methods and propose extensions of those methods. We limit our discussion to IRT-based methods. As a result, we avoid methods like the Mantel-Haenszel procedure [@holland1986differential], which muddies the waters by moving away from the IRT framework, and has been shown to perform no better — and in some cases worse — than IRT-based methods [@swaminathan1990detecting]. For simplicity, we focus on the Rasch model to isolate the fundamental issues in DIF detection. And, in search of a coherent framework, we sometimes edit names of existing methods. We recognize that others have done the same [e.g. @kopf2015framework] and that we risk contributing to the proliferation of names.

@@

In general, we denote the mean ability of the reference and focal group as $\mu^\text{ref}$ and $\mu^\text{foc}$, respectively. We follow the common and inconsequential practice of identifying the scale by setting $\mu^\text{ref} = 0$. The fundamental challenge for a DIF method is to first get an estimate for $\mu^\text{foc}$, $\hat\mu^\text{foc}$. How can group ability be disentangled from the possibility that items are biased? The most common approach is to identify a group of anchor items that are assumed to be DIF-free. These items are used to get $\hat\mu^\text{foc}$.

@@

We propose an extension of these methods, all-others-as-anchors-one-at-a-time (AOAA-OAT), which, to our knowledge (and surprise), has not previously been proposed. Like AOAA-AS, AOAA-OAT starts with AOAA, but only the single item exhibiting the most DIF, based on the $\chi^2$ test statistic, is removed from the anchor set. The process continues iteratively until no new items display DIF. AOAA-OAT and AOAA-AS are similar in that they are both iterative; the difference is that AOAA-OAT takes the more conservative approach of removing only one item in each iteration as opposed to all items that test positive for DIF. As a result, AOAA-OAT is less likely to arrive at the wrong conclusion because of artifical DIF.

>>>>

Psychometricians have long been in search of the perfect DIF detection method. In this section, we summarize existing DIF methods and propose extensions of those methods. We take as given that the IRT-based likelihood ratio test is the best way to test for DIF, which has been well demonstrated [@meade2012solving]. As a result, we avoid methods like the Mantel-Haenszel procedure [@holland1986differential], which muddies the waters by moving away from the IRT framework, and has been shown to perform no better — and in some cases worse — than IRT-based methods [@swaminathan1990detecting].

The unsolved and interesting problem is how to link groups in the common circumstance in which no a priori assumptions — about either relative group ability or which items contain bias — can be made. For simplicity, we focus on the Rasch model to isolate the fundamental issues in DIF detection. And, in search of a coherent framework, we sometimes edit names of existing methods. We recognize that others have done the same [e.g. @kopf2015framework] and that we risk contributing to the proliferation of names.

@@

In general, we denote the mean ability of the reference and focal group as $\mu^\text{ref}$ and $\mu^\text{foc}$, respectively. We follow the common and inconsequential practice of identifying the scale by setting $\mu^\text{ref} = 0$. The fundamental challenge for DIF methods is to disentangle estimating $\hat\mu^\text{foc}$ from estimating $\hat{b_j}^{\text{foc}} - \hat{b_j}^{\text{ref}}$. The most common approach is to identify a group of anchor items that are assumed to be DIF-free. These anchor items identify the model, thereby allowing for the estimation of $\hat\mu^\text{foc}$.

@@

We propose an extension of these methods, all-others-as-anchors-one-at-a-time (AOAA-OAT), which, to our knowledge (and surprise), has not previously been explicitly proposed. AOAA-OAT is inspired by @hagquist2017recent who, in general, assert that "items showing DIF initially
should not be resolved simultaneously but sequentially" [p. 6]. Like AOAA-AS, AOAA-OAT starts with AOAA, but only the single item exhibiting the most DIF, based on the $\chi^2$ test statistic, is removed from the anchor set. The process continues iteratively until no new items display DIF. AOAA-OAT and AOAA-AS are similar in that they are both iterative; the difference is that AOAA-OAT takes the more conservative approach of removing only one item in each iteration as opposed to all items that test positive for DIF. As a result, AOAA-OAT is less likely to arrive at the wrong conclusion because of artifical DIF.

### 2020-03-02 paper/paper_2b_anchor_points.Rmd
implementing non compensatory models which seems to be going well

```{r ginimilg, fig.cap = 'Selecting the anchor point by maximizing the Gini index.', out.width="70%"}

@@

Raju's area method measures the amount of DIF by calculating the area between the item characteristic curve, the function that maps ability to probability correct response, for each group [@raju1988area]:
```{r}
# {-\infty}^{\infty}
```
\begin{align}
    \text{Area Between Curves} = \int |\text{Pr}(y_j = 1| \theta, b_j^{\text{ref}}) - \text{Pr}(y_j = 1| \theta, b_j^{\text{foc}})|
\end{align}

Raju's area method, along with AOAA, have been cited as the two most commonly used IRT-based DIF detection methods [@magis2011generalized]. We see two weakness of Raju's area method: First, the item characteric curves still need to be linked by anchor items, which the method does not help to identify. Second, the area is unweighted so all values of $\theta$ matter equally, despite some being much more realistic than others.

>>>>

```{r ginimilg, fig.cap = 'Selecting the anchor point by maximizing the Gini index.', out.width="70%", warning = FALSE, message = FALSE}

@@

Raju's area method [@raju1988area] measures the amount of DIF by calculating the area between the item characteristic curve, the function that maps ability to probability correct response, for each group:
$$
\text{Area Between Curves} = \int |\text{Pr}(y_j = 1| \theta, b_j^{\text{ref}}) - \text{Pr}(y_j = 1| \theta, b_j^{\text{foc}})|
$$
Raju's area method, along with AOAA, have been cited as the two most commonly used IRT-based DIF detection methods [@magis2011generalized]. We see two weakness of Raju's area method: First, the item characteric curves still need to be linked by anchor items, which the method does not help to identify. Second, the area is unweighted so all values of $\theta$ matter equally, despite some being much more realistic than others.
